---
title: 'Monarchs Portfolio Assignment: Element 2'
author: "Matilde Elene"
date: "31st of July"
output:
  pdf_document: default
  html_document: default
bibliography: ../src/references.bib
csl: "../src/chicago-fullnote-bibliography.csl"
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      root.dir = ".")
```

```{r}
#loading packages 
library(pacman)
pacman::p_load(ggplot2, tidyverse, tibble, here,lubridate)
```


### 1. Load the monarchs
#### Setting directory paths: 
```{r}
#I created this dataset using create_data.Rmd in the src/ folder. Both the original dataset (danish_monarch.csv) and the cleaned dataset (monarchy_data.csv) are available in the data/ folder.
path_to_file <- "../data/monarchy_data.csv"
path_output <- "../out/"
```

***1.1:  Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)***

*List what is the separator:*

**Answer 1.1:** When opening the file in text edit, the dataset that I am loading has the delimiter [ , ]. In other words; the dataset is a standard comma-separated CSV.


### 2.  Create a `monarchs` object in R with the different functions below and inspect the different outputs.

-   `read.csv()`
-   `read_csv()`
-   `read.csv2()`
-   `read_csv2()`

```{r eval=TRUE, message=FALSE}
#trying different loading functions:
monarchs1 <- read.csv(path_to_file)
monarchs2 <- read_csv(path_to_file)
monarchs3 <- read.csv2(path_to_file)
monarchs4 <- read_csv2(path_to_file)
```

***2.1: Which of these functions is a `tidyverse` function? Read data with it below into a `monarchs` object***

**Answer 2.1:** The second and fourth functions (read_csv() and read_csv2()) are tidyverse functions from the readr package, where the first and third functions (read.csv() and read.csv2()) are base R functions. The main difference between read_csv() and read_csv2() (and similarly between read.csv() and read.csv2()) is their default delimiter. The first two expect (,), whereas the latter two expect semicolon-separated values (;) with commas used as decimal separators. In addition, the readr functions return tibbles, while the base R functions return data frames.\n

```{r eval=TRUE, message=FALSE}
#loading data with the correct function
monarchs <- read_csv(path_to_file)
```

***2.2: What is the result of running `class()` on the `monarchs` object created with a tidyverse function.***

```{r echo=TRUE, message=FALSE}
#cehcking class of dataframe. 
class_specs <- class(monarchs)
```

**Answer 2.2:** "Running class() on the monarchs object created with the tidyverse function (read_csv()) returns: "`r paste(class_specs, collapse = ", ")`." This indicates that the object is a tibble (tbl_df), which is the data structure returned by readr functions in the tidyverse."


***2.3: How many columns does the object have when created with these different functions?***

**Answer 2.3:** `read.csv()` and `read_csv()` each return 8 columns, whereas `read.csv2()` and `read_csv2()` each return 1 column. This outcome is expected because the dataset is comma-separated, while the latter two functions expect semicolon-separated values.

```{r echo=TRUE, message=FALSE}
rm(monarchs1, monarchs2, monarchs3, monarchs4) # clearing cache
```

***2.4: Show the dataset so that we can see how R interprets each column***

```{r eval = TRUE}
output2 <- dplyr::glimpse(monarchs) #using the dplyr package for a more tidy view of the dataset at hand.
```

**Answer 2.4:** The output of str() and glimpse() shows how R has interpreted each column, including its data type (e.g., character, integer, or numeric). This, however, is possible to override using the col_types(), as executed below:

```{r eval = TRUE}
#reloading the dataset with set columns, overriding the default 'guesses' from the read_csv file by using the col_types argument.)
monarchs <- read_csv(path_to_file,col_types = cols(date_birth = col_integer(), date_death = col_integer(), start_reign = col_integer(), end_reign = col_integer(), monarch_id = col_character()))
```

### 2. Calculate the duration of reign for all the monarchs in your table

You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

-   Is your data messy? Fix it before re-importing to R
-   Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`

Create a new column called `duration` in the monarchs dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r}
#renaming the columns in order to follow the wording of the exercise and creating a new column called 'duration' in the monarchs analysis dataset.
monarchs_analysis <- monarchs %>% 
  rename(endReign=end_reign, startReign=start_reign) %>%
  mutate(duration = endReign-startReign)

# inspecting the number of missing values in each column of the dataset
monarchs_analysis %>%  
  summarise(across(everything(), ~ sum(is.na(.))))

# as expected, duration has two missing values. Inspecting monarchs whose durations cannot be calculated
monarchs_analysis %>% 
  dplyr::filter(is.na(duration) == TRUE)

# For the purpose of the upcoming analysis these values are retained as NA, as their reign lengths cannot be calculated from the available information. They are excluded from analysis when a complete reign period is required. 
```

### 3. Calculate the average duration of reign for all rulers

***Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `monarchs` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!***

***3.1: You first need to know how to select the relevant `duration` column. What are your options?***

**Answer 3.1:** There are multiple options for selecting a specific column using R. Using the dplyr package, select() returns the chosen column as a one-column tibble (dataframe), and can be used through intermediate steps, (nested functions) or pipes. 

```{r}
#selecting through intermediate steps:(one-column tibble)
duration_tbl <- dplyr::select(monarchs_analysis, duration)
```

***3.2: Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.***

```{r}
#selecting through piping:
monarchs_analysis %>% 
  dplyr::select(duration) #this will be printed

# The class() function shows the object type.
cat("Class of monarchs_analysis['duration']:", class(monarchs_analysis["duration"]),"\n\n")
```
**Answer 3.2:** Using `select()` above returns an object of class `r paste(class(duration_tbl), collapse = ", ")`. The selected `duration` column is therefore stored as a one-column tibble.

Because 'duration'  is stored in a tibble, the summarise() function can be used to calculate its mean.
```{r}
avg_reign_tbl <- monarchs_analysis %>%
  summarise(mean_duration = mean(duration, na.rm = TRUE)) 
avg_reign_tbl
```
Alternatively, `$` extracts `duration` as a numeric vector:

```{r}
duration_vec <- monarchs_analysis$duration
class(duration_vec)

avg_reign_vec <- mean(duration_vec, na.rm = TRUE)
avg_reign_vec 
```
  
Both approaches produce an average reign duration of **`r round(avg_reign_vec, 2)` years**.
  
***3.3: Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.***

**Answer 3.3:** No coercion is required because `duration` is already stored as a numeric/integer variable.

***3.4: Remember to handle NAs: `mean(X, na.rm=TRUE)`***
**Answer 3.4:** For both calculations, `na.rm = TRUE` excludes missing (`NA`) values when calculating the mean. This is necessary because two observations have missing reign durations.



### 4. How many and which monarchs enjoyed a longer-than-average duration of reign?

***4.1: You have calculated the average duration above. Use it now to `filter()` the `duration` column in `monarchs` dataset. Display the result and also count the resulting rows with `count()`***

```{r fig.width=8, fig.height=4.5}
# filtering the dataset in order to view monarchs whose reigns were longer than average
over_avg_duration <- monarchs_analysis %>%
  dplyr::filter(duration > avg_reign_tbl$mean_duration)

#counting the remaining observations
n_total_over <- count(over_avg_duration)
```

**Answer 4.1:** After filtering for reign durations above the mean of **`r round(avg_reign_vec, 2)` years**, **`r n_total_over$n`** of the total 52 monarchs remain in the dataset.

```{r}
# Visualizing monarchs with longer-than-average reigns using ggplot()

# Defining the legend label for the average reign
avg_label <- paste0("Average reign (",avg_reign_vec," years)")

plot_over_average <- ggplot(
  over_avg_duration,
  aes(
    x = duration, 
    y = monarch_name, 
    fill = monarch_house)
  ) +
  geom_col(
    width = 0.6
  ) +
  geom_vline(
    aes(
      xintercept = avg_reign_vec,
      linetype = avg_label
      ),
    inherit.aes = FALSE,
      alpha = 0.6
    ) +
  scale_linetype_manual(
    name = NULL,
    values = setNames("dashed", avg_label)
    )+
  labs(
    fill = "House",
    x = "Duration of reign (years)",
    y = "Name of Monarch",
    title = paste0("Monarchs with reigns longer than average (n=", n_total_over$n," )"
    )
  ) +
  scale_fill_brewer(
    palette = "Dark2",
    labels = function(x) {
      x[x == "Schleswig-Holstein-Sonderburg-Glücksburg branch"] <- "Schleswig-Holstein-\nSonderburg-Glücksburg"
      x
    }
  ) +
  coord_cartesian(xlim = c(
    avg_reign_vec - 1,
    max(over_avg_duration$duration)
    )
    ) +
  guides(
    fill = guide_legend(order = 1),
    linetype = guide_legend(order = 2)
    )+
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8),
        legend.position = "right",
        legend.key.spacing.y = unit(0.2,"cm"),
        plot.title = element_text(face = "bold"))

plot_over_average

#saving plot to ../out
ggsave(
  filename = file.path(path_output, "over_average_reign.jpg"),
  plot = plot_over_average,
  width = 8.5,
  height = 6,
  dpi = 300 
)
```

### 5. How many days did the three longest-ruling monarchs rule?

***5.1: Sort monarchs by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function***
```{r, fig.width=8, fig.height=4.5}
#selecting the three longest-ruling monarchs by slicing and sorting by duration reign in descending order
three_longest <- over_avg_duration %>%
  arrange(desc(duration))%>%
  slice(1:3)

three_longest
```

**Answer 5.1:** The three longest-ruling monarchs are
**`r three_longest$monarch_name[1]`** (`r three_longest$duration[1]` years),
**`r three_longest$monarch_name[2]`** (`r three_longest$duration[2]` years), and
**`r three_longest$monarch_name[3]`** (`r three_longest$duration[3]` years).


***5.2: Use `mutate()` to create `Days` column where you calculate the total number of days they ruled***
-   *BONUS: consider the transition year (with 366 days) in your calculation!*

```{r, fig.width=8, fig.height=4.5}
#creating a "day" column in the original dataset, where the total number of days ruled are included (simply approximating 365 days a year (Gregorian Calendar)).
monarchs_analysis <- monarchs_analysis %>%
  mutate(days_of_reign = duration * 365) %>% #approximates each full year.
  arrange(desc(days_of_reign))

# Displaying the relevant columns first
monarchs_analysis %>%
  select(monarch_name, duration, days_of_reign, everything())
```

### 6. Challenge: Plot the monarchs' duration of reign through time

***6.1: Try to plot the duration of reign column in `ggplot` with `geom_point()` and `geom_smooth()`***
*- In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).*
*- Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis*
*- BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!*

**Answer 6.2:**
```{r warning=FALSE}
# Long term trend: duration of reign through time:
# creating the midyear column to represent each monarch's reign by its midpoint
monarchs_analysis <- monarchs_analysis %>% 
  mutate(midyear = startReign + (endReign-startReign)/2) # getting one data point rather than a ()

# creating gg-plot
plot_long_term <- ggplot(monarchs_analysis, 
                         aes(x=midyear, y=duration, color=monarch_house)) +
  geom_point(size=2, pch=8) + #closest I could get to a crown
  scale_x_continuous(
    breaks = seq(900,2100,by = 100)) + 
  scale_y_continuous(
    breaks = seq(0,60,by = 10 )
  ) +
  geom_smooth(
    aes(group = 1),
    color = "steelblue",
    fill = "steelblue",
    alpha=0.15,
    linewidth=0.4
    ) +
  scale_color_brewer(
    palette="Dark2",
    labels = function(x) {
      x[x == "Schleswig-Holstein-Sonderburg-Glücksburg branch"] <-
        "Schleswig-Holstein-\nSonderburg-Glücksburg"
      x[x == "House of Wittelsbach, Palatinate-Neumarkt branch"] <-
        "House of Wittelsbach,\nPalatinate-Neumarkt"
      x
      }
    ) +
  geom_hline(
    data=data.frame(avg=avg_reign_vec, label=avg_label),
    aes(yintercept = avg, linetype = label),
    inherit.aes = FALSE,
    alpha = 0.6
    ) +
  scale_linetype_manual(
    name = NULL,
    values = setNames("dashed", avg_label)
    ) +
  labs(
    x= "Midpoint",
    y="Duration of reign (years)",
    title = "Duration of reign through time",
    color="House"
  )+
  guides(
    color = guide_legend(order = 1),
    linetype = guide_legend(order = 2)
    )+
  theme_minimal() + #recreating the theme from the previous plot 
  theme(
    axis.title.y=element_text(size=10, margin = margin(r = 15)),
    axis.title.x = element_text(size=10, margin = margin(t = 15)),
    legend.position = "right",
    plot.title = element_text(hjust = 0.8, face="bold", size=14),
    legend.key.spacing.y = unit(0.2,"cm"))
    
plot_long_term

#saving plot to ./out
ggsave(
  filename = file.path(path_output, "long_term_plot.jpg"),
  plot = plot_long_term,
  width = 8.5,
  height = 6,
  dpi = 300 
)
```


***6.2: What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ?***

**Answer 6.2, Historical trends:**
The plot showing Duration of Danish Monarch's reign through time suggests a long-term increase in the duration of reign among Danish monarchs. Reigns were generally shorter and more variable during the earlier part of the period, particularly around the eleventh and twelfth centuries. Beginning with Gorm the Old, regarded as the first recognized king of Denmark, the earlier part of the series covers the Middle Ages, during which the duration of reign fluctuated considerably.^[@nationalmuseumofdenmark] From 1397–1523, the Kalmar Union brought the kingdoms of Denmark, Norway, and Sweden under a common monarch,^[@nationalmuseumofdenmark] while the plot indicates a gradual increase in reign duration around and following this period.

This development did not, however, correspond to an absence of warfare. Christian IV, for example, represents the longest reign in the dataset, despite ruling during a period of extensive military conflict.

The introduction of absolute monarchy under Frederik III in 1660 ^[@embassyofdenmark] represents another major change in Danish political organization, and the plot generally shows longer reigns during the subsequent period, although considerable variation remains.

The nineteenth century likewise combines relatively long reigns with political and military upheaval. Denmark entered the century on the losing side of the Napoleonic Wars, demonstrating that longer reigns cannot simply be interpreted as evidence for the disappearance of violence. Absolute monarchy was eventually abolished during the constitutional transition of 1848–1849. ^[@embassyofdenmark]

Overall, the increasing duration of reign is broadly assocciated with increasing political and institutional stability, but the historical examples show that long reigns could coexist with warfare and territorial conflict. Reign duration should therefore not be interpreted as a direct measure of historical violence.

::: {#footnotes}

:::

### Bibliography
::: {#refs}

:::