---
title: "Monarchs Portfolio Assignment: Element 2"
author: "Matilde Elene"
date: "31st of July"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      root.dir = ".")
```

```{r}
#loading packages 
library(pacman)
pacman::p_load(ggplot2, tidyverse, tibble, here,lubridate)
```

The task here is to load your monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the monarchs

Make sure to first create an `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `monarchs.csv` dataset.

```{r}
#I created this dataset using create_data.Rmd in the src/ folder. Both the original dataset (danish_monarch.csv) and the cleaned dataset (monarchy_data.csv) are available in the data/ folder.
path_to_file <- "../data/monarchy_data.csv"
path_output <- "../out/"
```

1.  Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the separator:

```{r, answer 1}
cat("When opening the file in text edit, the dataset that I am loading has the delimiter [ , ]. In other words; the dataset is a standard comma-separated CSV.")
```

2.  Create a `monarchs` object in R with the different functions below and inspect the different outputs.

-   `read.csv()`
-   `read_csv()`
-   `read.csv2()`
-   `read_csv2()`

```{r eval=TRUE}
monarchs1 <- read.csv(path_to_file)
monarchs2 <- read_csv(path_to_file)
monarchs3 <- read.csv2(path_to_file)
monarchs4 <- read_csv2(path_to_file)
```
Answer: 
1\. Which of these functions is a `tidyverse` function? Read data with it below into a `monarchs` object

2\. What is the result of running `class()` on the `monarchs` object created with a tidyverse function.

3\. How many columns does the object have when created with these different functions?

4\. Show the dataset so that we can see how R interprets each column

```{r eval = TRUE}
#Question 1: Which of these functions is a `tidyverse` function? 
cat("Answer 1: The second and fourth functions (read_csv() and read_csv2()) are tidyverse functions from the readr package, where the first and third functions (read.csv() and read.csv2()) are base R functions. The main difference between read_csv() and read_csv2() (and similarly between read.csv() and read.csv2()) is their default delimiter. The first two expect (,), whereas the latter two expect semicolon-separated values (;) with commas used as decimal separators. In addition, the readr functions return tibbles, while the base R functions return data frames.\n")

#loading data with the correct function
monarchs <- read_csv(path_to_file)

#question 2: 
class_specs <- class(monarchs)
cat("\nAnswer 2: Running class() on the monarchs object created with the tidyverse function (read_csv()) returns:",class_specs,".\nThis indicates that the object is a tibble (tbl_df), which is the data structure returned by readr functions in the tidyverse.\n")

#question 3:
cat("read.csv():", ncol(monarchs1), "columns\n", "read_csv():", ncol(monarchs2), "columns\n", "read.csv2():", ncol(monarchs3), "columns\n", "read_csv2():", ncol(monarchs4), "columns.\nThis outcome is expected, as the dataset I am loading is separated by commas. Therefore, the latter two functions will interpret these columns as one\n")

rm(monarchs1, monarchs2, monarchs3, monarchs4) # clearing cache

#question 4: showing the dataset so we can see how R interprets each column:
str(monarchs)

output2 <- dplyr::glimpse(monarchs) #using the dplyr package for a more tidy view of the dataset at hand.

cat("Answer 4: The output of str() and glimpse() shows how R has interpreted each column, including its data type (e.g., character, integer, or numeric). This, however, is possible to override using the col_types(), as executed below:\n")

#reloading the dataset with set columns, overriding the default 'fuesses' from the read_csv file by using the col_types argument.)
monarchs <- read_csv(path_to_file,col_types = cols(date_birth = col_integer(), date_death = col_integer(), start_reign = col_integer(), end_reign = col_integer(), monarch_id = col_character()))

```

## Calculate the duration of reign for all the monarchs in your table

You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

-   Is your data messy? Fix it before re-importing to R
-   Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`

Create a new column called `duration` in the monarchs dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r}
#renaming the columns in order to follow the wording of the exercise and creating a new column called 'duration' in the monarchs analysis dataset.
monarchs_analysis <- monarchs %>% 
  rename(endReign=end_reign, startReign=start_reign) %>%
  mutate(duration = endReign-startReign)

# inspecting the number of missing values in each column of the dataset
monarchs_analysis %>%  
  summarise(across(everything(), ~ sum(is.na(.))))

# as expected, duration has two missing values. Inspecting monarchs whose durations cannot be calculated
monarchs_analysis %>% 
  dplyr::filter(is.na(duration) == TRUE)

# For the purpose of the upcoming analysis these values are retained as NA, as their reign lengths cannot be calculated from the available information. They are excluded from analysis when a complete reign period is required. 
```

## Calculate the average duration of reign for all rulers

Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `monarchs` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

-   You first need to know how to select the relevant `duration` column. What are your options?
-   Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
-   Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
-   Remember to handle NAs: `mean(X, na.rm=TRUE)`

```{r}
# There are multiple options for selecting a specific column using R. Using the dplyr package, select() returns the chosen column as a one-column tibble (dataframe), and can be used through intermediate steps, (nested functions) or pipes. 

#selecting through intermediate steps:
duration_tbl <- dplyr::select(monarchs_analysis, duration)

#selecting through piping:
monarchs_analysis %>% 
  dplyr::select(duration) #this will be printed

# The class() function shows the object type.
cat("Class of monarchs_analysis['duration']:", class(monarchs_analysis["duration"]),"\n\n")

#Because 'duration'  is stored in a tibble, the summarise() function can be used to calculate its mean.
avg_reign_tbl <- monarchs_analysis %>%
  summarise(mean_duration = mean(duration, na.rm = TRUE)) 

cat("Average duration of reign (from tibble):",avg_reign_tbl$mean_duration,"\n\n")

# Since  mean() requires a numeric vector, the column is extracted as such.
duration_vec <- monarchs_analysis$duration
avg_reign_vec <- mean(duration_vec, na.rm = TRUE)

cat("Average duration of reign (from vector):",avg_reign_vec)

#for both calculations on the average reign duration, na.rm = TRUE automatically excludes missing (NA) values. 
```

## How many and which monarchs enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in `monarchs` dataset. Display the result and also count the resulting rows with `count()`

```{r fig.width=8, fig.height=4.5}
# filtering the dataset in order to view monarchs whose reigns were longer than average
over_avg_duration <- monarchs_analysis %>%
  dplyr::filter(duration > avg_reign_tbl$mean_duration)

str(over_avg_duration)
# from this specific filtering we've gone from 52 observations to 25. 

#counting the remaining observations
n_total_over <- count(over_avg_duration)
cat("Number of monarchs who have enjoyed a longer-than-average duration of reign:", n_total_over$n)

# Visualizing monarchs with longer-than-average reigns using ggplot()

# Defining the legend label for the average reign
avg_label <- paste0("Average reign (",avg_reign_vec," years)")

plot_over_average <- ggplot(
  over_avg_duration,
  aes(
    x = duration, 
    y = monarch_name, 
    fill = monarch_house)
  ) +
  geom_col(
    width = 0.6
  ) +
  geom_vline(
    aes(
      xintercept = avg_reign_vec,
      linetype = avg_label
      ),
    inherit.aes = FALSE,
      alpha = 0.6
    ) +
  scale_linetype_manual(
    name = NULL,
    values = setNames("dashed", avg_label)
    )+
  labs(
    fill = "House",
    x = "Duration of reign (years)",
    y = "Name of Monarch",
    title = paste0("Monarchs with reigns longer than average (n=", n_total_over$n," )"
    )
  ) +
  scale_fill_brewer(
    palette = "Dark2",
    labels = function(x) {
      x[x == "Schleswig-Holstein-Sonderburg-Glücksburg branch"] <- "Schleswig-Holstein-\nSonderburg-Glücksburg"
      x
    }
  ) +
  coord_cartesian(xlim = c(
    avg_reign_vec - 1,
    max(over_avg_duration$duration)
    )
    ) +
  guides(
    fill = guide_legend(order = 1),
    linetype = guide_legend(order = 2)
    )+
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8),
        legend.position = "right",
        legend.key.spacing.y = unit(0.2,"cm"),
        plot.title = element_text(face = "bold"))

plot_over_average

#saving plot to ../out
ggsave(
  filename = file.path(path_output, "over_average_reign.jpg"),
  plot = plot_over_average,
  width = 8.5,
  height = 6,
  dpi = 300     # High resolution for quality
)

```

## How many days did the three longest-ruling monarchs rule?

-   Sort monarchs by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
-   Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
-   BONUS: consider the transition year (with 366 days) in your calculation!

```{r, fig.width=8, fig.height=4.5}
#selecting the three longest-ruling monarchs by slicing and sorting by duration reign in descending order
three_longest <- over_avg_duration %>%
  arrange(desc(duration))%>%
  slice(1:3)

cat("The three longest-ruling monarchs are ", three_longest$monarch_name[1], " (", three_longest$duration[1], " years), ", three_longest$monarch_name[2], " (", three_longest$duration[2], " years), and ", three_longest$monarch_name[3], " (", three_longest$duration[3], " years).", sep = ""
)

#creating a "day" column in the original dataset, where the total number of days ruled are included (simply approximating 365 days a year (Gregorian Calendar)).
monarchs_analysis <- monarchs_analysis %>%
  mutate(days_of_reign = duration * 365) %>% #approximates each full year.
  arrange(desc(days_of_reign))

monarchs_analysis 
```

## Challenge: Plot the monarchs' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ?

-   Try to plot the duration of reign column in `ggplot` with `geom_point()` and `geom_smooth()`
-   In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
-   Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
-   BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r}
# Long term trend: duration of reign through time:

# creating the midyear column to represent each monarch's reign by its midpoint
monarchs_analysis <- monarchs_analysis %>% 
  mutate(midyear = startReign + (endReign-startReign)/2) # getting one data point rather than a ()

# creating gg-plot
plot_long_term <- ggplot(monarchs_analysis, 
                         aes(x=midyear, y=duration, color=monarch_house)) +
  geom_point(size=2, pch=8) + #closest I could get to a crown
  scale_x_continuous(
    breaks = seq(900,2100,by = 100)) + 
  scale_y_continuous(
    breaks = seq(0,60,by = 10 )
  ) +
  geom_smooth(
    aes(group = 1),
    color = "steelblue",
    fill = "steelblue",
    alpha=0.15,
    linewidth=0.4
    ) +
  scale_color_brewer(
    palette="Dark2",
    labels = function(x) {
      x[x == "Schleswig-Holstein-Sonderburg-Glücksburg branch"] <-
        "Schleswig-Holstein-\nSonderburg-Glücksburg"
      x[x == "House of Wittelsbach, Palatinate-Neumarkt branch"] <-
        "House of Wittelsbach,\nPalatinate-Neumarkt"
      x
      }
    ) +
  geom_hline(
    data=data.frame(avg=avg_reign_vec, label=avg_label),
    aes(yintercept = avg, linetype = label),
    inherit.aes = FALSE,
    alpha = 0.6
    ) +
  scale_linetype_manual(
    name = NULL,
    values = setNames("dashed", avg_label)
    ) +
  labs(
    x= "Midpoint",
    y="Duration of reign (years)",
    title = "Duration of reign through time",
    color="House"
  )+
  guides(
    color = guide_legend(order = 1),
    linetype = guide_legend(order = 2)
    )+
  theme_minimal() + #recreating the theme from the previous plot 
  theme(
    axis.title.y=element_text(size=10, margin = margin(r = 15)),
    axis.title.x = element_text(size=10, margin = margin(t = 15)),
    legend.position = "right",
    plot.title = element_text(hjust = 0.8, face="bold", size=14),
    legend.key.spacing.y = unit(0.2,"cm"))
    
plot_long_term

#saving plot to ./out
ggsave(
  filename = file.path(path_output, "long_term_plot.jpg"),
  plot = plot_long_term,
  width = 8.5,
  height = 6,
  dpi = 300     # High resolution for quality
)
```

```{r}
# historical trends: 
cat("From the plot (). 

For example, in ()

    King Gorm the old was the first recognized king of Denmark (DK). During the Middle Ages (app. )*. 
    
    From 1397-1523 the Kalmar Union tied together the Kingdoms of Norway, Sweden, and Denmark under one Monarch*
    
    Christian the 4th was one of the most renowned kings in Danish history - he ruled during the 17th and 18th century, in which the territories of Skåne, Halland, and Blekinge was lost to Sweden during this period*. 
    
    Because Denmark was on the losing side of the Nepoleonic Wars in the beginning of the 19th Century, ()*. 
    
    The Danish absolute monarchy was introduced by Frederik III with a coup in 1660. It was abolished in 1848 after the death of Christian VII**") 

#*footnote: Source is https://um.dk/litauen/en/about-lithuania/history/. **footnote: source is https://en.natmus.dk/historical-knowledge/denmark/absolute-monarchy-1660-1848/. 
```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output. Well done!
