---
title: "Monarchs Portfolio Assignment"
author: "Matilde Elene"
date: "31st of July"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The task here is to load your monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the monarchs

Make sure to first create an `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `monarchs.csv` dataset.

1.  Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the separator:\_\_\_\_\_\_\_

2.  Create a `kings` object in R with the different functions below and inspect the different outputs.

-   `read.csv()`
-   `read_csv()`
-   `read.csv2()`
-   `read_csv2()`

```{r eval=FALSE}
# FILL IN THE CODE BELOW and review the outputs
monarchs1 <- read.csv("data/_______")

monarchs2 <- read_csv("data/_______")

monarchs3 <- read.csv2("data/_______")

monarchs4 <- read_csv2( "data/_______")
```

Answer: 1. Which of these functions is a `tidyverse` function? Read data with it below into a `monarchs` object 2. What is the result of running `class()` on the `monarchs` object created with a tidyverse function. 3. How many columns does the object have when created with these different functions? 4. Show the dataset so that we can see how R interprets each column

```{r eval = FALSE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

monarchs <- _____

class(monarchs)


____(monarchs)


____(monarchs)


```

## Calculate the duration of reign for all the monarchs in your table

You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

-   Is your data messy? Fix it before re-importing to R
-   Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`

Create a new column called `duration` in the monarchs dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r}
# YOUR CODE
```

## Calculate the average duration of reign for all rulers

Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `monarchs` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

-   You first need to know how to select the relevant `duration` column. What are your options?
-   Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
-   Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
-   Remember to handle NAs: `mean(X, na.rm=TRUE)`

```{r}
# YOUR CODE
```

## How many and which monarchs enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in `monarchs` dataset. Display the result and also count the resulting rows with `count()`

```{r}
# YOUR CODE
```

## How many days did the three longest-ruling monarchs rule?

-   Sort monarchs by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
-   Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
-   BONUS: consider the transition year (with 366 days) in your calculation!

```{r}
# YOUR CODE
```

## Challenge: Plot the monarchs' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ?

-   Try to plot the duration of reign column in `ggplot` with `geom_point()` and `geom_smooth()`
-   In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
-   Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
-   BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r}
# YOUR CODE
```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output. Well done!
